Numerical experiment investigating the influence of drag on trajectory patterns of floating macroalgae

### 2021-05-05 thesis_chapter_3.Rmd
intro corrections...

---
title: 'Numerical experiment investigating the influence of drag on trajectory patterns of floating macroalgae'
author: 
- R Coppin^[Department of Biodiversity and Conservation Biology, University of the Western Cape, Private Bag X17, Bellville 7535, South Africa], C Rautenbach^[National Institute of Water and Atmospheric Research, 41 Market Place Viaduct Harbour, 1010 Auckland, Private Bag 99940, Newmarket, Auckland 1149, New Zealand], AJ Smit^[South African Environmental Observation Network, Elwandle Coastal Node, Port Elizabeth, South Africa]
output:
  word_document:
    reference_docx: markdown/wiley_template.docx
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document:
bibliography: references.bib
csl: markdown/apa.csl.txt
abstract: "**Abstract**: Ocean currents act as an important dispersal mechanism of material on the ocean surface, both natural and anthropogenic. Macroalgae are one of the most important natural dispersal vectors and are regarded as the 'tumble-weed' of the ocean. Despite many studies that exist on the topic, the relative role of wind and surface currents in influencing the trajectory of macroalgal dispersal is still unknown. Past studies focus on kelp-rafts, which can vary greatly in size, making it difficult to form a consenses as to the relative role of wind versus surface currents. In addition, past studies have not considered surface-area in relation to drag characteristics of floating macroalgae, both of which have been shown to play a role in the trajectory and accumulation of other flotsam. Advances in Lagrangian trajectory modeling have been rapid in recent years and allow the use of numerical experiments to investigate trajectories of flotsam in the ocean. This study aimed to shed light on the relative role of wind versus surface currents, and the role of drag in macrolagal trajectory and accumulation. We focused on solitary kelp individuals. This was achieved by comparing simulations of virtual kelp 'particles' which incorporate drag with that of purely Lagrangian particles. The results showed that the virtual kelp particles do follow the ocean currents in general, but that the inclusion of drag into the scenarios causes accumulation in eddys within the study domain. The results also point to the importance of sub-mesoscale features, such as eddys, as important surface accumulation zones for solitary floating macroalage."
---

@@

There is a range of objects---both natural and anthropogenic---floating in the ocean of which macroalgae are regarded as one of the most important passive dispersal mechanisms of marine taxa. Floating kelp acts as a passive dispersal mechanism for a range of marine taxa and is sometimes referred to as the 'tumble-weed' of the ocean [@edgar1987; @bushing1994; @norton1992; @helmuth1994; @holmquist1994; @smith2002; @McCormick2008]. Kelps also function as input of allochthonous materials to near and distant ecosystems and in this manner support the ecological function in a range of adjacent habitats [@cartraud2021; @colombini2003; @Filbee-Dexter2018-wp; @zuercher2019; @Krumhansl2012; @vanderklift2008; @bustamante1995].

Some macroalgal species are negatively buoyant and sink to the seafloor when detached from the substrate, while other species have air-filled pneumatocysts or stipes which allow the plants to reach the surface where light is more abundant. The positively buoyant pneumatocysts also cause plants to float to the surface when dislodged from the substratum. The structure and number of pneumatocysts varies between species. For example, species belonging to the genera *Macrocystis*, *Sargassum*, *Ascophyllum*, and *Fucus* have thalli with many small pneumatocysts, while other kelp species such as *Nereocystis luetkeana*, *Pelagophycus porra*, *Ecklonia radiata*, and *Ecklonia maxima* have single, large pneumatocysts [@dayton1985; @smith2002; @thiel2005; @graiff2016; @batista2018]. In some cases, the stipe itself may be air-filled, such as with *E. maxima* and *E. radiata*. Although there have been reports of floating species of Chlorophyta and Rhodophyta, Phaeophyceae species are the most reported forms of floating algae. This is most likely because the green and red species reported to float are not actually positively buoyant, but instead are kept at the surface by gas trapped in-between or in their thalli [@dromgoole1982; @back2000]. The giant kelp *Macrocystis pyrifera* [@helmuth1994; @kingsford1995; @hobday2000; @macaya2005; @graiff2016; @batista2018], the bull kelp *Durvillaea antarctica* [@collins2010; @tala2013 ; @tala2017; @saunders2014; @smith2002; @wichmann2012; @batista2018], and pelagic *Sargassum* species [@putman2020; @brooks2019; @putman2018; @brooks2018; @wang2019] have been the focus of much of the research regarding spatial and temporal dispersal patterns, the ecological role of rafting, marine connectivity and raft-time.

Past research points to macroalgal trajectories being largely determined by prevailing wind conditions and surface currents [@thiel2005; @hobday2000; @fraser2011; @rothausler2011; @rothausler2011]. Although ocean currents are regarded as the primary influence, the relative importance of wind versus surface current is still unknown although the role of wind has been recognized as important in several studies. For example, a study by @harrold1989 investigated the seasonal trajectories of radio-transmitter tagged *M. pyrifera* in nearshore Monterey Bay. The results showed that kelp rafts, with little surface area, exposed to the wind were largely driven by a combination of wind and surface currents; however, the relative importance of wind and was not clear. In addition, the trajectories of tagged kelp were more consistent with the formation of eddys during winter. Previous studies have also identified wind as an important mechanism of dispersal in wind dominated ocean systems. For example, in the sub-Antarctic latitudes the West Wind Drift causes continuous unidirectional surface flow and is regarded as an important potential mechanism for dispersal of floating bull kelp *Durvillaea antarctica*. Other studies have used genetic approaches to determine macroalgae raft trajectory characteristics by inferring source location from genetically similar populations [@nikula2013]. For example, a study by @fraser2011 on the rafting capabilities of *D. antarctica* used a combination of population genetics and relative age estimate of goose barnacles attached to the raft. The presence of goose barnacles suggests a long rafting time as these species have a slow growth rate. Genetic analyses showed these species can raft up to $\sim$ 390 km from their local origin. The authors suggested that wind and water-movement were the primary influences of trajectory; however, this was only inferred from the genetic results and local climatology data [@fraser2011]. Past research has shown that buoyancy and drag also play a role in determining the trajectory and rate of transport for surface floating material [@onink2019; @shen2019; @dasaro2018; @gates2018; @beron2016; @trinanes2016; @furnans2008]. The 'sail' area of an object floating at sea is the surface area exposed to the wind which results in air form drag, while the area of the object below the surface of the water is exposed to surface currents which result in hydrodynamic-form drag [@breivik2011; @griffin2017]. In addition, past research conducted by the maritime industry has shown that the size and shape of a vessel determine the relative importance of waves or wind as drivers of trajectory, as well as orientation of the object. If the length of the object is longer than the significant wave height then waves will be the primary driver of trajectory while the opposite is true for smaller objects where the effect of waves is regarded as negligible [@breivik2011; @griffin2017]. Drag is ultimately determined by the size and shape of the object which are properties that vary considerably with macroalgae species.

To accurately determine the trajectory of marine macroalgae, both water and wind drag need to be considered. Past research by @allen1999 and @breivik2011 have provided estimates of drag for various objects based on experimental work, which consisted of both direct and indirect methods of the Leeway model. However, this work does not consider biological material such as macroalgae. Although drag estimates do exist, these have been applied for macroalgae not detached from the substratum and are regarded as fixed-point estimates. Although various approaches have been used in the past to investigate floating macroalgal trajectories, very few studies have employed the use of Langrangian trajectory modeling. Furthermore, none of the existing studies using this modeling approach have considered macroalgal morphology cross-sectional area (*i.e.* shape) as an aspect of drag and ultimately trajectory. Studies by @brooks2019, @putman2018, and @putman2020 used Lagrangian approaches to investigate the effects of inertia, raft-size, and windage on macroalgae trajectory. @brooks2019 used a custom growth model to estimate changes in biomass and ultimately radial size, while a customized Hybrid Coordinate Ocean Model (HYCOM) for the trajectory simulations was used. The results showed that the trajectory of pelagic *Sargussum* was significantly influenced by inertia and the radial size of the rafts. Other work by @putman2020 investigated the effect of including windage in macroalgae trajectory simulations. The authors showed that including *ad hoc* windage factors greatly improved accuracy regarding entrainment or expulsion from eddys when compared to tracked *Sargassum* mats in the ocean. Langrangian trajectory modeling is clearly a useful tool; however, the various models available often assume the object to be a spherical particle, which is obviously not the case for both biological and anthropogenic forms of marine debris in nature. The radial size included by @brooks2019 is an improvement, however this approach does not consider the complex morphology on an individual level into account.

Past studies have focused on kelp-rafts, not solitary floating kelp, and therefore are not able to investigate the role of eddys and vortices in the accumulation of individual kelp plants at the surface. Gyres, eddys, and converging surface currents have been identified as important in the accumulation and dispersal of plastics and other marine debris [@onink2019; @dasaro2018; @brach2018]. Investigations into  the dispersal of marine debris have also included the possible effects of inertia in accumulation of ocean drifter instruments in mesoscale features. For example, a study by @beron2016 used a modeling approach to investigate the role of Eckman transport and inertia to explain the observed behavior of undrogued drifters in the ocean gyres. Undrogued drifter trajectory is affected by wind, unlike the drogue counter parts which makes the effects of wind negligible. The results showed that the combined effect inertia, not Eckman transport, played a significant role in the observed drifter behavior [@beron2016]. @beron2016 also noted that the presence of other flotsam, along with the undrogued drifters in ocean gyres, suggests that aspects of inertia also play a role in accumulation of other marine debris [@beron2016]. The approach by @beron2016 has also been improved upon by @beron2019 but also shows the significant role of inertia in flotsam trajectory and accumulation.

In this study, a numerical approach is used to shed light on the role of hydrodynamic and direct wind drag on macroalgal drift trajectory using Lagrangian based trajectory simulations. We aim to shed light on the 'behavior' of surface drifting macroalgae from a coastal environment into the open ocean, which will inform future characterization studies. We will achieve this by comparing the trajectory outputs of virtual passive particles (no drag characteristics added) with kelp particles (drag characteristics added) of varying degrees of hydrodynamic and wind drag exposure. The different degrees of wave and wind exposure are meant to reflect the different buoyancy characteristics, as well as identify the role of magnitude of the drag components on overall trajectory. The modeling of single kelp offers the opportunity to study the role of eddys and vortices in accumulation points on the surface.

@@

>>>>

---
title: 'Numerical experiment investigating the influence of drag on trajectory patterns of floating macroalgae'
author: 
- R Coppin^[Department of Biodiversity and Conservation Biology, University of the Western Cape, Private Bag X17, Bellville 7535, South Africa], C Rautenbach^[National Institute of Water and Atmospheric Research, 41 Market Place Viaduct Harbour, 1010 Auckland, Private Bag 99940, Newmarket, Auckland 1149, New Zealand], AJ Smit^[South African Environmental Observation Network, Elwandle Coastal Node, Port Elizabeth, South Africa]
output:
  word_document:
    reference_docx: markdown/wiley_template.docx
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document:
bibliography: references.bib
csl: markdown/apa.csl.txt
abstract: "**Abstract**: Ocean currents act as an important dispersal mechanism of material on the ocean surface, both natural and anthropogenic. Macroalgae are one of the most important natural dispersal vectors  of marine organisms and are regarded as the 'tumble-weed of the ocean'. Despite many studies that exist on the topic, the relative role of wind and surface currents in influencing the trajectory of macroalgal dispersal is still unknown. Past studies focus on kelp-rafts, which can vary greatly in size, making it difficult to form a consenses as to the relative role of wind versus surface currents. In addition, these studies have not considered surface-area in relation to drag characteristics of the macroalgae, both of which have been shown to play a role in the trajectory and accumulation of other flotsam. Advances in Lagrangian trajectory modeling have been rapid in recent years and allow the use of numerical experiments to investigate trajectories of flotsam in the ocean. This study aimed to shed light on the relative role of wind versus surface currents, and the role of drag in macrolagal trajectory and accumulation. We focused on solitary kelp individuals. This was achieved by comparing simulations of virtual kelp 'particles' which incorporate drag with that of purely Lagrangian particles. The results showed that the virtual kelp particles do follow the ocean currents in general, but that the inclusion of drag into the scenarios causes accumulation in eddys within the study domain. The results also point to the importance of sub-mesoscale features, such as eddys, as important surface accumulation zones for solitary floating macroalage."
---

@@

A range of natural and anthropogenic objects floats in the ocean, and macroalgae are regarded as one of the most important of these passive dispersal mechanisms of marine taxa. Floating kelps act as passive dispersal mechanisms for a range of marine taxa and is sometimes referred to as the ‘tumble-weed of the ocean’ [@edgar1987; @bushing1994; @norton1992; @helmuth1994; @holmquist1994; @smith2002; @McCormick2008]. Kelps also function as input of allochthonous materials to near and distant ecosystems and in this manner support the ecological function in a range of adjacent habitats [@cartraud2021; @colombini2003; @Filbee-Dexter2018-wp; @zuercher2019; @Krumhansl2012; @vanderklift2008; @bustamante1995].

Some macroalgal species are negatively buoyant and sink to the seafloor when detached from the substrate, while other species have air-filled pneumatocysts or stipes which allow the plants to reach the surface where light is more abundant. The positively buoyant pneumatocysts also cause plants to float to the surface when dislodged from the substratum. The structure and number of pneumatocysts varies among species. For example, species belonging to the genera *Macrocystis*, *Sargassum*, *Ascophyllum*, and *Fucus* have thalli with many small pneumatocysts, while other kelp species such as *Nereocystis luetkeana*, *Pelagophycus porra*, *Ecklonia radiata*, and *Ecklonia maxima* have single, large pneumatocysts [@dayton1985; @smith2002; @thiel2005; @graiff2016; @batista2018]. In some cases, the stipe itself may be air-filled, such as with *E. maxima* and *E. radiata*. Although there have been reports of floating species of Chlorophyta and Rhodophyta, Phaeophyceae species are the most reported forms of floating algae. This is most likely because the green and red species reported to float are not actually positively buoyant, but instead are kept at the surface by gas trapped in-between or in their thalli [@dromgoole1982; @back2000]. The giant kelp *Macrocystis pyrifera* [@helmuth1994; @kingsford1995; @hobday2000; @macaya2005; @graiff2016; @batista2018], the bull kelp *Durvillaea antarctica* [@collins2010; @tala2013 ; @tala2017; @saunders2014; @smith2002; @wichmann2012; @batista2018], and pelagic *Sargassum* species [@putman2020; @brooks2019; @putman2018; @brooks2018; @wang2019] have been the focus of much of the research regarding spatial and temporal dispersal patterns, the ecological role of rafting, marine connectivity and raft-time.

Past research points to macroalgal trajectories being largely determined by prevailing wind conditions and surface currents [@thiel2005; @hobday2000; @fraser2011; @rothausler2011; @rothausler2011]. Although ocean currents are regarded as the primary influence, the relative importance of wind versus surface current is still unknown although the role of wind has been recognized as important in several studies. For example, a study by @harrold1989 investigated the seasonal trajectories of radio-transmitter tagged *M. pyrifera* in nearshore Monterey Bay. The results showed that kelp rafts, with little surface area, exposed to the wind were largely driven by a combination of wind and surface currents; however, the relative importance of wind and was not clear. In addition, the trajectories of tagged kelp were more consistent with the formation of eddys during winter. Studies have also identified wind as an important mechanism of dispersal in wind dominated ocean systems. For example, in the sub-Antarctic latitudes the West Wind Drift causes continuous unidirectional surface flow and is regarded as an important potential mechanism for dispersal of floating bull kelp *Durvillaea antarctica*. Other studies have used genetic approaches to determine macroalgae raft trajectory characteristics by inferring source location from genetically similar populations [@nikula2013]. For example, a study by @fraser2011 on the rafting capabilities of *D. antarctica* used a combination of population genetics and relative age estimate of goose barnacles attached to the raft. The presence of goose barnacles suggests a long rafting time as these species have a slow growth rate. Genetic analyses showed these species can raft up to $\sim$ 390 km from their local origin. The authors suggested that wind and water-movement were the primary influences of trajectory; however, this was only inferred from the genetic results and local climatology data [@fraser2011]. Research has shown that buoyancy and drag also play a role in determining the trajectory and rate of transport for surface floating material [@onink2019; @shen2019; @dasaro2018; @gates2018; @beron2016; @trinanes2016; @furnans2008]. The 'sail' area of an object floating at sea is the surface area exposed to the wind which results in air form drag, while the area of the object below the surface of the water is exposed to surface currents which result in hydrodynamic-form drag [@breivik2011; @griffin2017]. Work by the maritime industry has shown that the size and shape of a vessel determine the relative importance of waves or wind as drivers of trajectory, as well as orientation of the object. If the length of the object is longer than the significant wave height then waves will be the primary driver of trajectory while the opposite is true for smaller objects where the effect of waves is regarded as negligible [@breivik2011; @griffin2017]. Drag is ultimately determined by the size and shape of the object which are properties that vary considerably with macroalgae species.

To accurately determine the trajectory of marine macroalgae, both water and wind drag need to be considered. @allen1999 and @breivik2011 have provided estimates of drag for various objects based on experimental work, which consisted of both direct and indirect methods of the Leeway model. However, this work did not consider biological material such as macroalgae. Although drag estimates do exist, these have been applied for macroalgae not detached from the substratum and are regarded as fixed-point estimates [@Denny1997; @Denny2002a; @Denny2002b; @Gaylord1997-ow; @Gaylord2008-qv]. Although various approaches have been used in the past to investigate floating macroalgal trajectories, very few studies have employed the use of Langrangian trajectory modeling. Furthermore, none of these studies using this modeling approach have considered macroalgal morphology cross-sectional area (*i.e.* shape) as an aspect of drag and ultimately trajectory. @brooks2019, @putman2018, and @putman2020 used Lagrangian approaches to investigate the effects of inertia, raft-size, and windage on macroalgae trajectory. For example, @brooks2019 used a custom growth model to estimate changes in biomass and ultimately radial size, while a customized Hybrid Coordinate Ocean Model (HYCOM) for the trajectory simulations was used. Their results showed that the trajectory of pelagic *Sargussum* was significantly influenced by inertia and the radial size of the rafts. @putman2020 investigated the effect of including windage in macroalgae trajectory simulations and showed that including *ad hoc* windage factors greatly improved accuracy regarding entrainment or expulsion from eddys when compared to tracked *Sargassum* mats in the ocean. Langrangian trajectory modeling is clearly a useful tool; however, the various models available often assume the object to be a spherical particle, which is obviously not the case for both biological and anthropogenic forms of marine debris in nature. The radial size included by @brooks2019 is an improvement but this approach does not consider the complex morphology on an individual level into account.

Studies have focused on kelp-rafts, not solitary floating kelp, and therefore are not able to investigate the role of eddys and vortices in the accumulation of individual kelp plants at the surface. Gyres, eddys, and converging surface currents have been identified as important in the accumulation and dispersal of plastics and other marine debris [@onink2019; @dasaro2018; @brach2018]. Investigations into  the dispersal of marine debris have also included the possible effects of inertia in accumulation of ocean drifter instruments in mesoscale features. For example, @beron2016 used a modeling approach to investigate the role of Eckman transport and inertia to explain the observed behavior of undrogued drifters in the ocean gyres. Undrogued drifter trajectory is affected by wind, unlike the drogue counter parts which makes the effects of wind negligible. The results showed that the combined effect inertia, not Eckman transport, played a significant role in the observed drifter behavior [@beron2016]. These authors also noted that the presence of other flotsam, along with the undrogued drifters in ocean gyres, suggests that aspects of inertia also play a role in accumulation of other marine debris [@beron2016]. The approach by @beron2016 has also been improved upon by @beron2019 but they also highlight the significant role of inertia in flotsam trajectory and accumulation.

In this study, a numerical approach is used to shed light on the role of hydrodynamic and direct wind drag on macroalgal drift trajectory using Lagrangian based trajectory simulations. We aim to determine the 'behavior' of surface drifting macroalgae from a coastal environment to the open ocean, which will inform future characterization studies. We will achieve this by comparing the trajectory outputs of virtual passive particles (no drag characteristics added) with kelp particles (drag characteristics added) of varying degrees of hydrodynamic and wind drag exposure. The different degrees of wave and wind exposure are meant to reflect the different buoyancy characteristics, as well as identify the role of magnitude of the drag components on overall trajectory. The modeling of single kelp offers the opportunity to study the role of eddys and vortices in accumulation points on the surface.

@@

<div id="refs"></div>
